The role of advection versus diffusion in intraventricular transport

### 2025-03-12 paper/main.tex
update IC adv-diff, time-stepping, mesh

\subsection{Image-based computational geometries of zebrafish brain ventricles}
To model CSF flow and transport in the zebrafish brain ventricles, we
generated a 3D surface representation of the ventricular walls using
confocal imaging data of a zebrafish embryo injected
intraventricularly with a 70 kDa dye at 2
dpf~\cite{Olstad2019CiliaryDevelopment}. At 2 dpf, the ventricular
system consists mainly of 3 cavities: the telencephalic (anterior),
the di-/mesencephalic (middle) and the rhombencephalic (posterior)
ventricles, which are connected by interventricular ducts. From the
surface geometry, we constructed a volumetric mesh of the ventricles
using fTetWild~\cite{Hu2020FastWild}, consisting of 132 134
tetrahedral cells with a maximal (minimal) edge length of 14.9 \textmu
m (4.68 \textmu m) (\Crefpanels{fig:fig1}{d}). Parts of the
ventricular surfaces were marked to be lined with motile cilia
(\Crefpanels{fig:fig1}{d}, magenta
markers)~\cite{Olstad2019CiliaryDevelopment}.

@@

\subsection{Computational model of solute transport after photoconversion}
We simulate transport of photoconverted proteins within the brain ventricles by
modeling transport of a solute with concentration $c(\xx, t)$ for $\xx\in\Omega$ and
time $t > 0$ by the advection-diffusion equation
\begin{equation}
    \pdifft{c} + \nabla\cdot\JJ = 0 \quad \text{in } \Omega,
    \label{eq:adv_diff_strong}
\end{equation}
where the total concentration flux $\JJ$ is assumed to consist of an advective and a diffusive flux:
\begin{equation*}
    \JJ = c\uu - D\nabla c .
\end{equation*}
Here, $\uu$ is the CSF velocity field, governed by the Stokes equations (\Cref{eq:stokes_eqs}),
and $D$ is the molecular diffusion coefficient of the solute.
Initially, we set $c(\xx, 0)=0$ for all $\xx\in\Omega$.
The ventricular walls are assumed to be impermeable (admitting \emph{no flux}: $\JJ\cdot\nn=0$),
except at the anterior/posterior boundary $\Gp$.
At the anterior/posterior boundary, we permit the solute concentration to be advected into
and out of the domain in a manner that ensures conservation of mass
(see~\Cref{subsec:adv_diff_eq_weak_form} for details).

To computationally represent protein photoconversion in a region
$\Omega_{\mathrm{pc}} \subset \Omega$ (corresponding to ROI 1),
we prescribe a given time-dependent value to the solute concentration in this region. 
Mimicking the experimental protocol, we consider the dorsal part of the
diencephalic ventricle as the subdomain $\Omega_{\mathrm{pc}}$, and set
\begin{equation}
  c(\xx, t)=\frac{\log{(1+t/a)}}{\log{(1+T/a)}},
  \quad \xx \in \Omega_{\mathrm{pc}}, \quad t > 0.
\label{eq:photoconversion_curve}
\end{equation}
The value $a=65$ was chosen to fit the fluorescence intensity curve
observed in the physical photoconversion experiments.

@@

\subsection{Numerical approximation of the advection-diffusion equation}
The advection-diffusion equation (\Cref{eq:adv_diff_strong}) is discretized
with a first-order implicit Euler method in time, and a first-order,
symmetric interior penalty DG
method in space~\cite{Arnold1982AnElements}, chosen due to its favorable
conservation properties.
To accurately approximate the solute concentrations even
in the presence of strong advection (high Péclet numbers),
we use an upwind scheme for the velocity when calculating the
advective flux~\cite{Patankar2018NumericalFlow}.
As for the flow equations, we use a timestep size of
$\Delta t = 0.023$ s. We simulate transport for 1900 cardiac cycles,
with a final time of simulation $T = 856$ s.
A DG penalty parameter $\alpha = 25$ was chosen to ensure stability of the method.

>>>>

\subsection{Image-based computational geometries of zebrafish brain ventricles}
To model CSF flow and transport in the zebrafish brain ventricles, we
generated a 3D surface representation of the ventricular walls using
confocal imaging data of a zebrafish embryo injected
intraventricularly with a 70 kDa dye at 2
dpf~\cite{Olstad2019CiliaryDevelopment}. At 2 dpf, the ventricular
system consists mainly of 3 cavities: the telencephalic (anterior),
the di-/mesencephalic (middle) and the rhombencephalic (posterior)
ventricles, which are connected by interventricular ducts. From the
surface geometry, we constructed a volumetric mesh of the ventricles
by initially using fTetWild~\cite{Hu2020FastWild} and subsequently
local refinement functionality in~\cite{dolfinx2023preprint}. The mesh
consists of 141 512 tetrahedral cells with a maximal (minimal) edge length of 14.9 \textmu
m (3.1 \textmu m) (\Crefpanels{fig:fig1}{d}). Parts of the
ventricular surfaces were marked to be lined with motile cilia
(\Crefpanels{fig:fig1}{d}, magenta
markers)~\cite{Olstad2019CiliaryDevelopment}.

@@

\subsection{Computational model of solute transport after photoconversion}
We simulate transport of photoconverted proteins within the brain ventricles by
modeling transport of a solute with concentration $c(\xx, t)$ for $\xx\in\Omega$ and
time $t > 0$ by the advection-diffusion equation
\begin{equation}
    \pdifft{c} + \nabla\cdot\JJ = 0 \quad \text{in } \Omega,
    \label{eq:adv_diff_strong}
\end{equation}
where the total concentration flux $\JJ$ is assumed to consist of an advective and a diffusive flux:
\begin{equation*}
    \JJ = c\uu - D\nabla c .
\end{equation*}
Here, $\uu$ is the CSF velocity field, governed by the Stokes equations (\Cref{eq:stokes_eqs}),
and $D$ is the molecular diffusion coefficient of the solute. 
The ventricular walls are assumed to be impermeable (admitting \emph{no flux}: $\JJ\cdot\nn=0$),
except at the anterior/posterior boundary $\Gp$.
At the anterior/posterior boundary, we permit the solute concentration to be advected into
and out of the domain in a manner that ensures conservation of mass
(see~\Cref{subsec:adv_diff_eq_weak_form} for details).

To computationally represent protein photoconversion in a region
$\Omega_{\mathrm{pc}} \subset \Omega$ (corresponding to ROI 1),
we prescribe a given time-dependent value to the solute concentration in this region. 
Mimicking the experimental protocol, we consider the dorsal part of the
diencephalic ventricle as the subdomain $\Omega_{\mathrm{pc}}$, and set
\begin{equation}
  c(\xx, t)=\frac{\log{(1+t/a)}}{\log{(1+T/a)}},
  \quad \xx \in \Omega_{\mathrm{pc}}, \quad t > 0.
\label{eq:photoconversion_curve}
\end{equation}
The value $a=65$ was chosen to fit the fluorescence intensity curve
observed in the physical photoconversion experiments. 
To impose a smooth initial condition, we initially solve the diffusion equation
\begin{equation*}
    \pdifft{c} = \Tilde{D}\nabla^2 c
\end{equation*}
with an increased diffusion coefficient $\Tilde{D}=D\times 10^{5}$ for two timesteps.
The solutions are normalized by their respective maximum values, and subsequently
multiplied with the value of \Cref{eq:photoconversion_curve} at the times $t=\Delta t$
and $t=2\Delta t$, respectively.
The resulting concentration fields are used as initial conditions for $c(\xx, 0)$ and
$c(\xx, \Delta t)$ for all $\xx\in\Omega$, where $\Delta t$
is the timestep. The two initial conditions are required because of the
two-step temporal discretization scheme employed (see~\Cref{subsec:num_approx_adv_diff} for more details.)

@@

\subsection{Numerical approximation of the advection-diffusion equation}\label{subsec:num_approx_adv_diff}
The advection-diffusion equation (\Cref{eq:adv_diff_strong}) is discretized
with with a second-order, symmetric interior penalty DG
method in space~\cite{Arnold1982AnElements}, chosen due to its favorable
conservation properties. In time, the equation is discretized with the second-order backward
difference scheme BDF2.~\cite{Volker2016}
To accurately approximate the solute concentrations even
in the presence of strong advection (high Péclet numbers),
we use an upwind scheme for the velocity when calculating the
advective flux~\cite{Patankar2018NumericalFlow}.
As for the flow equations, we use a timestep size of
$\Delta t = 0.023$ s. We simulate transport for 1900 cardiac cycles,
with a final time of simulation $T = 856$ s.
A DG penalty parameter $\alpha = 50$ was chosen to ensure stability of the method.
